---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ROCnGO

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/ROCnGO)](https://CRAN.R-project.org/package=ROCnGO)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

## Overview

ROCnGO provides a set of tools to study a classifier performance by using ROC curve based analysis. Package may address tasks in these type of analysis such as:

* Evaluating global classifier performance.
* Evaluating local classifier performance when a high specificity or sensitivity is required, by using different indexes that provide: 
  * Better interpretation of local performance.
  * Better power of discrimination between classifiers with similar performance.
* Evaluating performance on several classifier simultaneously.
* Plot whole, or specific regions, of ROC curves.

## Installation

Last version of ROCnGO can be installed from its [GitHub](https://github.com/pabloPNC/ROCnGO) repository with:

``` r
# install.packages("devtools")
devtools::install_github("pabloPNC/ROCnGO")
```

## Usage

```{r, warning=FALSE}
library(ROCnGO)

# Iris subset
iris_subset <- iris[iris$Species != "versicolor", ]

# Select Species = "virginica" as the condition of interest
iris_subset$Species <- relevel(iris_subset$Species, "virginica")

# Summarize a predictor over high sensitivity region
summarize_predictor(
  iris_subset,
  predictor = Sepal.Length,
  response = Species,
  threshold = 0.9,
  ratio = "tpr"
)

# Summarize several predictors simultaneously
summarize_dataset(
  iris_subset,
  predictor = c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width),
  response = Species,
  threshold = 0.9,
  ratio = "tpr"
)

# Plot ROC curve of classifiers
plot_roc_curve(iris_subset, predictor = Sepal.Length, response = Species) +
  add_roc_curve(iris_subset, predictor = Petal.Length, response = Species) +
  add_roc_points(iris_subset, predictor = Sepal.Width, response = Species) +
  add_chance_line()
```


